Emulating running tzar in R
===========================

The emulation functionality explained in this document is aimed at allowing 
you to run and debug jobs in an interactive environment that replicates some of 
the most useful elements of tzar.  

Debugging and developing R code that runs under tzar can sometimes be difficult 
because you lack interactive access to console I/O and to debugging 
tools when a tzar job is running in the background.  You _can_ emulate this 
process interactively by running an R program outside of tzar 
and mocking up some of tzar's useful services, such as 
building the parameters list from the yaml file.  However, this 
mocking up is time-consuming and not necessarily reliable or identical to tzar.  

The code described in this document provides an interactive and very low 
overhead alternative in the form of a tzar emulator that also guarantees a perfect match to the parameter lists and directory structures that tzar would provide.  The idea 
behind the emulator is very simple:

1) You embed a call to the emulator near the 
start of your own project code and you source your code in the same way that you 
would normally run a standalone R program.  

2) The emulation function then 
runs tzar externally, but only up to the 
point where your model.R file calls your project code.  This means that tzar 
just decodes the yaml file and creates the tzar output directory.  

3) It will then save the location of that 
output directory to a scratch file and return to the emulator function in 
your project's source code.  There, the emulator retrieves the output directory 
location from the scratch file.  

4) The emulator then goes to that output directory 
and sources the variables.R file where tzar always caches the parameters list that it 
decoded from the yaml file.

From that 
point on, your project code runs just like it did under a normal local tzar run 
because it now has the same parameters list and the same output directory.  

This document is divided into 2 parts.  The first part walks through each of 
the steps that are involved in an emulation run of tzar.  The second part 
looks at running the example code in the example-tzar-emulator-R project.


Steps in the emulation
---


### Loading the code

First, you need to load the code and variables that drive the emulation.

```{r}
source ('emulateRunningUnderTzar.R')
```

### The controlling variable: emulateRunningUnderTzar

There is a boolean variable in that file that controls whether emulation is 
done or not.  That variable is called emulateRunningUnderTzar.  

If you set it to TRUE, then you can run your code in a tzar emulation.  This 
means that you just run your R program as you would normally run it, e.g., 
in the R command window or in RStudio.

If you set it to FALSE, then you can run tzar as you normally would.

I have initially set it to be TRUE for this explanation.  If the following 
command displays it as FALSE, then go back and edit emulateRunningUnderTzar.R 
so that emulateRunningUnderTzar=TRUE and rerun this explanation so that the 
emulateRunningUnderTzar.R file is sourced again.  

If you don't do this, then you may get a message like:
"## Error: object 'parameters' not found" further down in this explanation 
when it tries to find the parameters list.

```{r}
cat ("emulateRunningUnderTzar = ", emulateRunningUnderTzar)
```

###  Embedding the call to the emulation in your code

Once that is set up, you would insert the following commands near the start of 
your source code that you want to run in tzar emulation mode.  

The emulator needs to use a slightly different command for Windows than for 
other operating systems, so you either need to ask for the name of the OS 
here or have already asked for it somewhere before you get to the call below 
to emulateRunningTzar().  

The call to emulateRunningTzar() will run tzar on your project so that tzar 
will build the list of parameters from the yaml file and create the output 
directory.  emulateRunningTzar() will then retrieve the list of parameters 
and return that list.  

If emulateRunningUnderTzar=FALSE, then it would not 
have called emulateRunningTzar() and tzar would have loaded the parameters 
list and called your project's source code itself.  In that case, all console 
output would have gone to the tzar log file and you would not have been able 
to have any interaction with the program while it was running (e.g., using the 
browser() command or any of the debugging tools in RStudio).

```{r}
current.os = sessionInfo()$R.version$os

if (emulateRunningUnderTzar)
    parameters = emulateRunningTzar (current.os, tzarEmulation_scratchFileName)
```

###  Executing your program's normal activities

Now that the emulation setup has run, you should have the same paraemeters list 
in the parameters variable that tzar would have given you in a normal tzar run.
Note that this includes all of the wildcard substitutions, etc.  

At this point, you can do whatever you would normally do in your code.  As a 
very simple example, I'll just print out the values of the parameter list.

```{r}
cat ("\n\n========  parameters = \n")
print (parameters)
cat ("\n\n========  END parameters = \n")
```

###  Cleaning up after the emulation

Finally, once your program has finished doing whatever it normally does, 
you can call a routine to clean up a couple of things that the emulation 
either changed or left behind.  

For example, it creates a scratch file in 
your project directory that is no longer useful, so the 
cleanUpAfterTzarEmulation() routine will delete that file.  You could delete 
that file yourself if you did not call the cleanup routine or you could leave 
it alone and it would be overwritten the next time the emulator was run.  
Cleaning it up seems like it would usually be the best idea so that there's 
not a lot of meaningless junk around.

One other thing that cleaning up does is it allows you to rename the tzar 
output directory's extension to indicate that the run was an emulation rather 
than a normal tzar run.  

While tzar is running normally, it attaches 
".inProgress" to the output directory name, e.g., 
"1_default_scenario.inProgress".  When the run completes successfully, tzar 
removes that extension to leave just 1_default_scenario".  If the run fails, 
then tzar attaches a ".failed" extension, as in "1_default_scenario.failed".  

To designate the end of an emulation run, the cleanup routine will append 
an extension of your choice (including no extension), with a default value of 
".completedTzarEmulation", e.g., "1_default_scenario.completedTzarEmulation".  
This value can be changed by changing the setting of the 
tzarEmulation_completedDirExtension variable in emulateRunningUnderTzar.R.

After the cleanup is finished, there should be no scratch file left in your 
project area and the tzar output directory should have your designated 
extension (or the default extension if you didn't change it).

One thing to note is that your code can call the cleanup function regardless 
of whether your run succeeded or failed, so having the extension on the 
directory does not imply that your run succeeded.  However, because it's just 
a variable in the program, you could get fancy and reset the extension value 
yourself before calling the cleanup.  

The primary reason why an extension is added at all is to make it clear which 
tzar outputs are real tzar runs vs. emulations.  If this is not done, then it 
could be confusing to come back to some old runs and find that their log 
files are nearly empty in spite of the fact that tzar appeared to have run 
successfully.
```{r}
if (emulateRunningUnderTzar)
    {
    cat ("\n\nCleaning up after running emulation...\n\n")
    cleanUpAfterTzarEmulation (parameters)
    }
```

The example emulation project
---

The example-tzar-emulator-R project contains a very simple example of how you 
could put all this together.  It consists of 4 primary files:

- model.R
    - the mainline code expected and called first by tzar
- project.yaml
    - the yaml file that tzar expects as the source of parameter settings
- example-tzar-emulator-R.R
    - the source code file that would be where your project's code lives
- emulateRunningUnderTzar.R
    - the source code and parameter file for the emulator
    
To try this code under emulation, make sure that 
emulateRunningUnderTzar.R has the 3 most important variables set correctly for 
your situation:

- *emulateRunningUnderTzar* set to TRUE
- *projectPath* set to the full path to the directory of your project, e.g.,
    - "/Users/Bill/D/rdv-framework/projects/example-tzar-emulator-R/"
- *tzarJarPath* set to the full path to the tzar jar file to be executed, e.g., 
    - "/Users/Bill/D/rdv-framework/tzar.jar"

and then just source example-tzar-emulator-R.R:

```{r}
source ('example-tzar-emulator-R.R')
```

The output from sourcing the example code lists the parameters loaded for the 
project.  One element of that list is called $fullOutputDirWithSlash and it 
shows where the run was done, but it has a ".inProgress" extension.  The 
cleanup routine will have changed that to a ".completedTzarEmulation" extension 
and you can look in that area to see the results of the run.

One thing that you will notice is that if you open the logging.log file, unlike 
a normal tzar run, the log file  
will not contain the same things that were echoed to the screen here.  For 
example, one line of the output here says 

"========  parameters IN emulation = "

Because the emulated tzar only ran far enough to load up parameters from the 
yaml file and to create the output directory, the log file won't show anything 
after that.  Instead, you will already have had access to it in your normal 
output console.  

However, anything that your program did outside of the console should be just 
the same as it would have been in a tzar run.  For example, if you had created 
subdirectories for pieces of your output or you had written output files or 
plots, tose should all be exactly where they would have been in a normal 
tzar run.